---
title: Maps
---

```{r}
library(tidyverse)
library(maps)
library(mapproj)
library(ggthemes)
```

# Create Original Map

`geom_map()` matches each row's `map_id` to a region in a separate `map` data set. `expand_limits()` is required, or ggplot won't know the x/y range to show. `coord_map()` gives a nicer projection than the default, and `theme_map()` removes the axes.

```{r}
states <- map_data("state")
ggplot(states) +
  geom_map(aes(map_id = region), map = states, color="gold2", fill="navyblue") +
  expand_limits(x = states$long, y = states$lat) +
  coord_map() +
  theme_map()
```

## Your turn

Edit the code so that each state is a different color

::: callout-tip
`states` already has a `region` column — you can map it to `fill` the same way it's mapped to `map_id`
:::

```{r}
states <- map_data("state")
ggplot(states) +
  geom_map(aes(map_id = region), map = states, color="gold2", fill="navyblue") +
  expand_limits(x = states$long, y = states$lat) +
  coord_map() +
  theme_map()
```

# Your turn: ACS data

## Load Data

```{r}
acs_state_data = read_csv("acs_state_data_2022_5y.csv") 
acs_state_data$state = tolower(acs_state_data$NAME)
```

Variable information:

-   `med_age`: median age
-   `med_income`: median age
-   `race_X`: estimated population of (self-reported) race
-   `born_in_state`: estimated population who were born in state
-   `X_age_married`: average age at first marriage for self-reported male and female respondents
-   `hs_diploma`, `associate_degree`, `prof_degree`, etc.: estimated number with a high school diploma, associate's degree, professional degree, etc.
-   `internet_X`: estimated number of people who have internet services

## ACS Data Starter Map

**Important note:** this is the same `geom_map()` you just used, now with real data. `acs_state_data` only needs a column with the state name (`state`). `geom_map()` looks up the matching shape in `map = states` for you.

```{r}
ggplot(acs_state_data) + 
  geom_map(
    aes(map_id = state, fill = NAME),
    map = states
  ) +
  expand_limits(x = states$long, y = states$lat) +
  coord_map() +
  theme_map()
```

### Your turn 1

Comment out the `expand_limits` line using `#`. What happened? 

### Your turn 2

Edit your chloropleth map by:

- Choosing a different variable in `acs_state_data` to map to `fill`
- Choosing a different color scale
- Updating the title, axis labels, and legend title

::: callout-tip

:::

# Try it: Proportional symbol maps

So far we've used `geom_map()` to fill in regions. But sometimes you want to show *where specific things are located*, with size showing how big they are. To do this, we add a second layer with `geom_point()`, using  latitude/longitude coordinates.

## Load Data

The `{maps}` package (already loaded) includes `us.cities`, a data set of US cities and their populations (as of 2006). We'll filter to cities with a population over 100,000, and exclude Alaska and Hawaii so the map isn't stretched out.

```{r}
data(us.cities)

big_cities <- us.cities |>
  filter(!country.etc %in% c("AK", "HI"), pop > 100000)
```

- `name`: city and state
- `pop`: population
- `lat`, `long`: the city's location

## Starter map: base + points

::: callout-tip
Notice that `geom_map()` and `geom_point()` each have their *own* `data`. The base map uses `states`; the points use `big_cities`. This is the same pattern you'll need on your homework.
:::

```{r}
#| error: true
ggplot(states) +
  geom_map(aes(map_id = region), 
               map = states, 
               color = "white", 
               fill = "gray90") +
  geom_point(data = big_cities, 
              aes(x = ___, y = ___, size = ___), 
              color = "navyblue", 
              alpha = 0.6) +
  scale_size(range = c(1, 10), labels = scales::comma) +
  expand_limits(x = states$long, y = states$lat) +
  coord_map() +
  theme_map()
```

### Your turn

- Fill in the `___` to overlay the points on the map
- Adjust the point `size` range, `alpha`, and color until you're happy with how the map looks

```{r}
# your turn

```
