---
title: "Draft: Proportional symbol maps (for 05-maps.qmd)"
---

```{r}
library(tidyverse)
library(maps)
library(mapproj)
library(ggthemes)
```

```{r}
#| echo: false
states <- map_data("state")
```

# Proportional symbol maps

So far we've used `geom_map()` to fill in *regions* — states colored by a variable. But sometimes you want to show *where specific things are located*, with size showing how big they are, like the wind turbines you'll map on your homework. For that, we add a second layer: `geom_point()`, using real latitude/longitude coordinates, layered right on top of a `geom_map()` base.

## Load Data

The `{maps}` package (already loaded) includes `us.cities`, a data set of US cities and their populations. We'll filter to cities with a population over 500,000, and exclude Alaska and Hawaii so the map isn't stretched out.

```{r}
data(us.cities)

big_cities <- us.cities |>
  filter(!country.etc %in% c("AK", "HI"), pop > 500000)
```

- `name`: city and state
- `pop`: population
- `lat`, `long`: the city's location

## Starter map: base + points

::: callout-tip
Notice that `geom_map()` and `geom_point()` each have their *own* `data`. The base map uses `states`; the points use `big_cities`. This is the same pattern you'll need on your homework.
:::

```{r}
ggplot(states) +
  geom_map(aes(map_id = region), map = states, color = "white", fill = "gray90") +
  geom_point(data = big_cities, aes(x = long, y = lat, size = pop), color = "navyblue", alpha = 0.6) +
  scale_size(range = c(1, 10), labels = scales::comma) +
  expand_limits(x = states$long, y = states$lat) +
  coord_map() +
  theme_map()
```

### Your turn

- Adjust the point `size` range, `alpha`, and color until you're happy with how the map looks

```{r}
# your turn

```
