---
title: "Stat 220 — Package & Environment Check"
subtitle: "Fall 2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: paged
    code-tools: true
execute:
  error: true      # keep rendering even if a chunk errors — we WANT to see failures
  warning: true    # show warnings (deprecation notices matter after a server update)
  message: true
  freeze: false
---

## What this file is for

The Maize RStudio server was updated, which usually means updated R package
versions. Render this document (**Render** button, or
`quarto render playground/package-check.qmd` in the Terminal) to confirm the
toolkit used in Stat 220 still works.

**How to read the results**

- The **availability table** lists every package the course uses and whether it
  is installed, with its version. Anything in the **"MISSING"** callout needs to
  be installed before the term starts.
- The **functional checks** run a small piece of real code from each topic area.
  Each line prints `PASS` or `FAIL: <message>`. A `FAIL` means a function the
  course relies on changed or broke.
- Warnings are printed on purpose. A new **deprecation warning** on code we teach
  is worth a look even if it still `PASS`es.

If something fails, see [If a check fails](#if-a-check-fails) at the bottom.

```{r}
#| label: setup

# Load the everyday packages up front. `require()` (not `library()`) so a missing
# one doesn't abort this chunk — the availability table below reports it clearly.
for (p in c("tidyverse", "scales", "lubridate")) {
  suppressPackageStartupMessages(require(p, character.only = TRUE))
}

# Toggle the network-dependent checks (APIs, scraping, remote database).
# Set to FALSE if you're off the campus network / VPN and only want the
# local package checks.
run_network_checks <- TRUE

# Helper: run an expression, print PASS / FAIL, never stop the render.
ok <- function(label, expr) {
  result <- tryCatch({
    force(expr)
    "PASS"
  }, error = function(e) paste0("FAIL: ", conditionMessage(e)))
  cat(sprintf("%-6s %s\n", result, label))
  invisible(result)
}
```

## R session

```{r}
#| label: session
cat(R.version.string, "\n")
cat("Platform:      ", R.version$platform, "\n")
cat("Running under: ", sessionInfo()$running, "\n")
cat("Pandoc:        ", as.character(rmarkdown::pandoc_version()), "\n")
cat("Quarto:        ", as.character(quarto::quarto_version()), "\n")
cat(".libPaths():\n"); cat(paste0("  ", .libPaths()), sep = "\n")
```

## Package availability

```{r}
#| label: availability

pkg_groups <- list(
  "Core tidyverse" = c(
    "tidyverse", "ggplot2", "dplyr", "tidyr", "readr", "purrr",
    "tibble", "stringr", "forcats", "lubridate"
  ),
  "Plotting add-ons" = c(
    "ggthemes", "patchwork", "scales", "gghighlight", "cowplot",
    "colorspace", "hrbrthemes", "ggdist", "viridisLite", "RColorBrewer"
  ),
  "Colour-vision / accessibility" = c("colorBlindness", "colorblindr", "thematic"),
  "Data sets" = c(
    "palmerpenguins", "nycflights23", "gapminder", "babynames",
    "unvotes", "mdsr", "survivoR", "bakeoff", "maps"
  ),
  "Spatial / maps" = c("sf", "mapproj", "ggmap", "leaflet", "tidycensus"),
  "Text" = c("tidytext", "textdata", "stopwords", "countrycode"),
  "Web / APIs / scraping" = c(
    "rvest", "httr2", "robotstxt", "googlesheets4", "gtrendsR", "nytimes"
  ),
  "Databases" = c("DBI", "dbplyr", "duckdb", "RMariaDB"),
  "Interactive / tables" = c("shiny", "flexdashboard", "bslib", "plotly", "DT", "gt"),
  "Reporting / infrastructure" = c(
    "knitr", "rmarkdown", "quarto", "tinytex", "here", "janitor",
    "kableExtra", "flextable", "fontawesome"
  ),
  "Instructor / slide authoring" = c("countdown", "flipbookr", "ymlthis", "tictoc"),
  "Git / dev tools" = c("usethis", "gitcreds", "remotes", "devtools")
)

availability <- do.call(rbind, lapply(names(pkg_groups), function(g) {
  pkgs <- pkg_groups[[g]]
  data.frame(
    group     = g,
    package   = pkgs,
    installed = vapply(pkgs, function(p) requireNamespace(p, quietly = TRUE), logical(1)),
    version   = vapply(pkgs, function(p) {
      tryCatch(as.character(packageVersion(p)), error = function(e) NA_character_)
    }, character(1)),
    row.names = NULL,
    stringsAsFactors = FALSE
  )
}))

knitr::kable(availability)
```

```{r}
#| label: missing-summary
missing <- availability$package[!availability$installed]
if (length(missing) == 0) {
  cat("All listed packages are installed.\n")
} else {
  cat("MISSING —", length(missing), "package(s) not installed:\n\n")
  cat(paste0("  ", missing), sep = "\n")
  cat('\n\nInstall the CRAN ones with:\n\n  install.packages(c(',
      paste(sprintf('"%s"', setdiff(missing, c("colorblindr", "nytimes", "flipbookr"))), collapse = ", "),
      '))\n', sep = "")
  cat("\nGitHub-only packages (need `remotes`):\n",
      '  remotes::install_github("clauswilke/colorblindr")\n',
      '  remotes::install_github("EvaMaeRey/flipbookr")\n',
      '  remotes::install_github("mkearney/nytimes")   # legacy, API demo only\n', sep = "")
}
```

Notes on a few of these:

- **`textdata`** downloads sentiment lexicons on first use (interactive prompt).
  This file only checks that it loads.
- **`ggmap`** / **`tidycensus`** need an API key to fetch data; only the load is
  checked here.
- **`nytimes`**, **`colorblindr`**, **`flipbookr`** are GitHub-only.

## Functional checks

### Core tidyverse

```{r}
#| label: check-core
ok("readr::read_csv parses an inline CSV", {
  d <- readr::read_csv(I("x,y\n1,a\n2,b\n"), show_col_types = FALSE)
  stopifnot(nrow(d) == 2, is.numeric(d$x))
})
ok("dplyr group_by |> summarise |> arrange", {
  mpg |>
    dplyr::group_by(manufacturer) |>
    dplyr::summarise(n = dplyr::n(), mean_hwy = mean(hwy), .groups = "drop") |>
    dplyr::arrange(dplyr::desc(n))
})
ok("dplyr::mutate + across + case_when", {
  mpg |> dplyr::mutate(
    dplyr::across(c(displ, hwy), \(x) scales::rescale(x)),
    size = dplyr::case_when(cyl <= 4 ~ "small", cyl <= 6 ~ "mid", TRUE ~ "big")
  )
})
ok("tidyr pivot_longer -> pivot_wider round trip", {
  wide <- tidyr::pivot_wider(
    tidyr::pivot_longer(head(iris), -Species, names_to = "k", values_to = "v"),
    names_from = k, values_from = v
  )
  stopifnot(ncol(wide) == ncol(head(iris)))
})
ok("purrr::map_dbl", {
  stopifnot(purrr::map_dbl(1:5, \(x) x^2) == c(1, 4, 9, 16, 25))
})
ok("stringr regex (str_detect / str_extract / str_replace_all)", {
  x <- c("STAT 220", "MATH 241", "seminar")
  stopifnot(
    stringr::str_detect(x, "[0-9]{3}") == c(TRUE, TRUE, FALSE),
    stringr::str_extract("due 2026-09-14 at 5pm", "[0-9]{4}-[0-9]{2}-[0-9]{2}") == "2026-09-14",
    stringr::str_replace_all("a1b2c3", "[0-9]", "") == "abc"
  )
})
ok("forcats::fct_reorder / fct_lump", {
  f <- forcats::fct_lump(factor(rep(letters[1:5], times = c(10, 5, 3, 1, 1))), n = 2)
  forcats::fct_reorder(f, seq_along(f), .fun = length)
})
ok("lubridate parse + component + arithmetic", {
  d <- lubridate::ymd("2026-09-14")
  stopifnot(
    lubridate::wday(d, label = TRUE) == "Mon",
    d + lubridate::weeks(10) == lubridate::ymd("2026-11-23")
  )
})
```

### Plots (shown inline — eyeball these)

```{r}
#| label: check-plots
#| fig-width: 8
#| fig-height: 3
library(ggplot2)

p1 <- ggplot(mpg, aes(displ, hwy, colour = drv)) +
  geom_point(alpha = .7) +
  scale_colour_viridis_d() +
  ggthemes::theme_fivethirtyeight() +
  labs(title = "ggplot2 + scales + ggthemes")

p2 <- ggplot(dplyr::count(mpg, class), aes(reorder(class, n), n)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::label_comma()) +
  coord_flip() +
  labs(title = "scales labels", x = NULL, y = NULL)

# patchwork side-by-side
p1 + p2
```

```{r}
#| label: check-plots-2
#| fig-width: 8
#| fig-height: 3.5
ok("gghighlight", {
  print(
    ggplot(mpg, aes(displ, hwy, colour = class)) + geom_point() +
      gghighlight::gghighlight(class == "suv")
  )
})
ok("cowplot::plot_grid", {
  print(cowplot::plot_grid(p1, p2, labels = c("A", "B")))
})
ok("ggdist half-eye", {
  print(
    ggplot(mpg, aes(factor(cyl), hwy)) +
      ggdist::stat_halfeye()
  )
})
ok("colorspace::scale_colour_discrete_qualitative", {
  print(
    ggplot(mpg, aes(displ, hwy, colour = drv)) + geom_point() +
      colorspace::scale_colour_discrete_qualitative()
  )
})
```

### Course data packages

```{r}
#| label: check-data
ok("palmerpenguins::penguins", {
  stopifnot(nrow(palmerpenguins::penguins) > 300,
            "species" %in% names(palmerpenguins::penguins))
})
ok("nycflights23::flights", {
  f <- nycflights23::flights
  stopifnot(nrow(f) > 4e5, "carrier" %in% names(f))
})
ok("gapminder::gapminder", stopifnot(nrow(gapminder::gapminder) == 1704))
ok("babynames::babynames", stopifnot("prop" %in% names(babynames::babynames)))
ok("unvotes::un_votes", stopifnot(nrow(unvotes::un_votes) > 0))
ok("mdsr (package loads)", stopifnot(requireNamespace("mdsr", quietly = TRUE)))
ok("survivoR::castaways", stopifnot(nrow(survivoR::castaways) > 0))
ok("bakeoff::bakers / ratings", stopifnot(nrow(bakeoff::ratings) > 0))
ok("maps::map_data('state')", stopifnot(nrow(ggplot2::map_data("state")) > 1000))
```

### Maps / spatial

```{r}
#| label: check-spatial
#| fig-width: 7
#| fig-height: 4
ok("sf::st_read + geom_sf (bundled nc.shp)", {
  nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
  print(ggplot(nc) + geom_sf(aes(fill = AREA)) + theme_void() +
          labs(title = "sf + geom_sf"))
})
ok("mapproj via coord_map('albers')", {
  states <- ggplot2::map_data("state")
  print(
    ggplot(states, aes(long, lat, group = group)) +
      geom_polygon(fill = "grey90", colour = "white") +
      coord_map("albers", lat0 = 39, lat1 = 45) +
      theme_void()
  )
})
ok("leaflet widget builds", {
  m <- leaflet::leaflet() |> leaflet::addTiles() |>
    leaflet::setView(-93.15, 44.46, zoom = 13)   # Northfield
  stopifnot(inherits(m, "leaflet"))
})
ok("ggmap loads (API key needed for tiles)",
   stopifnot(requireNamespace("ggmap", quietly = TRUE)))
ok("tidycensus loads (API key needed for data)",
   stopifnot(requireNamespace("tidycensus", quietly = TRUE)))
```

```{r}
#| label: leaflet-show
leaflet::leaflet() |> leaflet::addTiles() |>
  leaflet::addMarkers(-93.1536, 44.4619, popup = "Carleton College")
```

### Text analysis

```{r}
#| label: check-text
ok("tidytext::unnest_tokens + stop-word anti_join", {
  txt <- tibble::tibble(
    doc = c(1, 2),
    text = c("The quick brown fox", "jumps over the lazy dog")
  )
  tokens <- txt |>
    tidytext::unnest_tokens(word, text) |>
    dplyr::anti_join(tidytext::get_stopwords(), by = "word")
  stopifnot(nrow(tokens) >= 5, !"the" %in% tokens$word)
})
ok("stopwords::stopwords('en')", stopifnot(length(stopwords::stopwords("en")) > 100))
ok("countrycode::countrycode", {
  stopifnot(countrycode::countrycode("US", "iso2c", "country.name") == "United States")
})
ok("textdata loads (lexicons download on first use)",
   stopifnot(requireNamespace("textdata", quietly = TRUE)))
```

### Reading spreadsheets

```{r}
#| label: check-import
ok("readxl reads a bundled sample .xlsx", {
  xl <- readxl::read_excel(readxl::readxl_example("clippy.xlsx"))
  stopifnot(nrow(xl) > 0)
})
ok("googlesheets4 loads + gs4_deauth()", {
  googlesheets4::gs4_deauth()
  TRUE
})
```

### Web / APIs / scraping

```{r}
#| label: check-scrape-local
ok("rvest::read_html + html_table (no network)", {
  h <- rvest::read_html(
    "<table><tr><th>course</th><th>n</th></tr><tr><td>STAT220</td><td>25</td></tr></table>"
  )
  tbl <- rvest::html_table(h)[[1]]
  stopifnot(nrow(tbl) == 1, tbl$course == "STAT220")
})
ok("httr2 builds a request + req_dry_run (no network)", {
  req <- httr2::request("https://example.com/api") |>
    httr2::req_url_query(subject = "STAT", term = "26FA") |>
    httr2::req_headers(Accept = "application/json")
  httr2::req_dry_run(req, quiet = TRUE)
})
```

```{r}
#| label: check-scrape-network
#| eval: !expr run_network_checks
cat("Network checks (set run_network_checks <- FALSE to skip):\n\n")
ok("robotstxt::paths_allowed()", {
  res <- robotstxt::paths_allowed("https://www.google.com/search")
  stopifnot(is.logical(res))
})
ok("rvest fetches a live page", {
  pg <- rvest::read_html("https://www.r-project.org")
  stopifnot(length(rvest::html_elements(pg, "a")) > 0)
})
ok("httr2 performs a live request", {
  resp <- httr2::request("https://httpbin.org/get") |>
    httr2::req_timeout(15) |>
    httr2::req_perform()
  stopifnot(httr2::resp_status(resp) == 200)
})
ok("gtrendsR::gtrends", {
  g <- gtrendsR::gtrends("r programming", time = "today 1-m")
  stopifnot(is.list(g))
})
```

### Databases

```{r}
#| label: check-db-local
ok("DBI + duckdb + dbplyr: write, query, collect (all local)", {
  con <- DBI::dbConnect(duckdb::duckdb())
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE), add = TRUE)
  DBI::dbWriteTable(con, "mtcars", mtcars)
  out <- dplyr::tbl(con, "mtcars") |>
    dplyr::group_by(cyl) |>
    dplyr::summarise(mpg = mean(mpg, na.rm = TRUE), .groups = "drop") |>
    dplyr::collect()
  stopifnot(nrow(out) == 3)
})
```

```{r}
#| label: check-db-network
#| eval: !expr run_network_checks
ok("mdsr::dbConnect_scidb('airlines') — needs campus network + RMariaDB", {
  db <- mdsr::dbConnect_scidb("airlines")
  on.exit(try(DBI::dbDisconnect(db), silent = TRUE), add = TRUE)
  tbls <- DBI::dbListTables(db)
  stopifnot(length(tbls) > 0)
})
```

### Interactive output (htmlwidgets / tables)

```{r}
#| label: check-widgets
ok("shiny UI + server objects build (not launched)", {
  ui <- shiny::fluidPage(
    shiny::sliderInput("n", "n", 1, 100, 30),
    shiny::plotOutput("hist")
  )
  server <- function(input, output, session) {
    output$hist <- shiny::renderPlot(hist(rnorm(input$n)))
  }
  stopifnot(inherits(ui, c("shiny.tag", "shiny.tag.list")), is.function(server))
})
ok("bslib::bs_theme()", stopifnot(inherits(bslib::bs_theme(), "bs_theme")))
ok("flexdashboard loads", stopifnot(requireNamespace("flexdashboard", quietly = TRUE)))
```

```{r}
#| label: gt-table
mpg |>
  dplyr::count(class, drv) |>
  tidyr::pivot_wider(names_from = drv, values_from = n, values_fill = 0) |>
  gt::gt() |>
  gt::tab_header(title = "gt table renders")
```

```{r}
#| label: dt-table
DT::datatable(head(mpg, 20), options = list(pageLength = 5))
```

```{r}
#| label: plotly-widget
#| fig-width: 8
#| fig-height: 4
plotly::ggplotly(
  ggplot(palmerpenguins::penguins, aes(bill_length_mm, body_mass_g, colour = species)) +
    geom_point()
)
```

### Reporting toolchain

```{r}
#| label: check-reporting
ok("here::here()", stopifnot(dir.exists(here::here())))
ok("janitor::clean_names()", {
  d <- janitor::clean_names(data.frame(`A B` = 1, `c.d` = 2, check.names = FALSE))
  stopifnot(identical(names(d), c("a_b", "c_d")))
})
ok("knitr::kable()", stopifnot(nchar(knitr::kable(head(mtcars))[1]) > 0))
ok("kableExtra::kbl()", stopifnot(requireNamespace("kableExtra", quietly = TRUE)))
ok("flextable::flextable()", stopifnot(inherits(flextable::flextable(head(mtcars)), "flextable")))
ok("fontawesome::fa()", stopifnot(grepl("svg", fontawesome::fa("r-project"))))
ok("tinytex installed?", {
  cat("\n  tinytex::is_tinytex() = ", tinytex::is_tinytex(), "\n", sep = "")
  TRUE
})
```

## PDF rendering

Portfolio "sketch drafts" and lab quizzes render to PDF, so the LaTeX toolchain
matters too. `tinytex::is_tinytex()` above should be `TRUE`. To exercise the full
path, run this in the Terminal:

```bash
quarto render playground/package-check.qmd --to pdf
```

If it fails on missing `.sty` files, `tinytex` normally auto-installs them; if it
is broken after the server update, `tinytex::reinstall_tinytex()` fixes it.

## If a check fails

| Symptom | Fix |
|---|---|
| Package in the **MISSING** list | `install.packages("<pkg>")`, or the `remotes::install_github()` line printed above for GitHub-only ones |
| `FAIL: could not find function "..."` | Function renamed/removed in an update — check the package's NEWS/changelog and update the course material that uses it |
| New **deprecation warning** on taught code | Still works, but plan to update the example before students see it |
| `duckdb` / `sf` fail to load with a system-library error | System dependency missing on the server — send the error to the sysadmin |
| Network checks fail but local checks pass | Likely VPN/firewall, not a package problem — re-run on the campus network, or set `run_network_checks <- FALSE` |
| PDF render fails | `tinytex::reinstall_tinytex()` |

```{r}
#| label: sessioninfo
sessionInfo()
```
