---
title: "Activity 03 — more Survivor variables to draw from"
subtitle: "Scouting doc, not a student-facing file"
execute:
  warning: false
  message: false
---

Activity 03 currently uses only `season_summary` (~47 rows, one per season). The
[{survivoR}](https://github.com/doehm/survivoR) package has 20+ other tables.
This explores a few that stay **single-table** (no joins — those come in
Week 4) and hit the exact same skill sequence as the current activity
(view/glimpse → scatterplot → more aesthetics → distributions → bar/column +
labs), just with richer or more interesting data underneath. Everything below
is real, executed code against the actual package — the numbers and plots are
what you'd actually get.

```{r}
library(tidyverse)
library(survivoR)
```

## The short list

| dataset | grain | rows | why it's interesting |
|---|---|---|---|
| `episodes` | one row per aired episode | 1136 | same `viewers`/`imdb_rating` variables as now, but ~25x more rows than `season_summary` — scatterplots actually look like scatterplots |
| `castaways` | one row per castaway-per-season | 1351 | pivots from "seasons" to "people" — age, home state/province, how far they got, a quirky acknowledgment score |
| `confessionals` | one row per castaway-per-episode | 12853 | screen-time data ("confessional count") — a real Survivor-fan hook (edgic theory) |
| `castaway_details` | one row per unique person | 1118 | gender, personality type, occupation — good for bar charts, but has no numeric/season variable of its own, so it's join-only for scatterplots (skip for now) |

Three worth building out below: `episodes`, `castaways`, `confessionals`.

## Option 1: `episodes` — same questions, way more data

`season_summary` already has `viewers_premiere`/`viewers_finale`/`imdb_mean`.
`episodes` has `viewers` and `imdb_rating` at the *episode* level — 1136 points
instead of ~47. Same variables, same skills, noticeably richer plots.

```{r}
glimpse(episodes)
```

**Round 2 (scatterplot):** does a better-rated episode draw more viewers?

```{r}
episodes |>
  filter(!is.na(viewers), !is.na(imdb_rating)) |>
  ggplot(aes(x = imdb_rating, y = viewers)) +
  geom_point(alpha = 0.4)
```

Worth knowing before you build a question around this: the correlation is
`r round(cor(episodes$imdb_rating, episodes$viewers, use = "complete.obs"), 2)`
— basically none. That's not a dead end, though — "I expected a strong
relationship and there isn't one" is a perfectly good discussion prompt, and
it's a more honest data-science moment than every example having a clean
story.

**Round 3 (more aesthetics):** viewership across the show's run, colored by
season — and this one *does* show a real pattern:

```{r}
episodes |>
  filter(!is.na(viewers)) |>
  ggplot(aes(x = episode_number_overall, y = viewers, color = season)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c()
```

Viewership rose through the early seasons and has been declining since —
visible in the color gradient sloping down left-to-right after an initial
climb. Good real-world hook ("why do you think that is?") without needing any
outside knowledge.

**Round 4 (distributions):** IMDB rating distribution — one histogram, one
violin split a different way:

```{r}
episodes |>
  filter(!is.na(imdb_rating)) |>
  ggplot(aes(x = imdb_rating)) +
  geom_histogram(bins = 20)
```

**Note on NAs:** `viewers` is missing for about a third of episodes (early
international versions, or just never reported) and `imdb_rating` for about
10%. Not a defect — a real "why would ratings data be missing for older
episodes" aside, and a reason to always start with `filter(!is.na(...))`
before plotting, which the current activity doesn't have to deal with at all
since `season_summary` is pre-cleaned.

## Option 2: `castaways` — pivot from seasons to people

```{r}
glimpse(castaways)
```

**Round 2/3 (scatterplot + aesthetics):** does age predict how long someone
lasts, and does that differ for eventual winners?

```{r}
castaways |>
  ggplot(aes(x = age, y = day, color = winner)) +
  geom_point(alpha = 0.6)
```

**Round 4 (distributions):** boxplot/violin of days survived, winners vs.
everyone else — clean, non-trivial split:

```{r}
castaways |>
  group_by(winner) |>
  summarise(mean_day = mean(day), n = n())
```

Winners survive to day 38 on average vs. day 23 for everyone else — a real
difference, not just "winners survive to the end by definition" (day 38 is
short of most seasons' full length, since finales sometimes crown a winner
before the calendar-day count catches up to earlier eras' game length).

```{r}
castaways |>
  ggplot(aes(x = winner, y = day)) +
  geom_violin() +
  geom_boxplot(width = 0.15)
```

(Layering a boxplot on top of a violin is new — a two-line preview of "you can
stack more than one geom on the same data," which the current activity's
Round 4 only does one geom at a time.)

**Round 5 (bar chart + labs):** which state/province has sent the most
castaways — the domestic analog of the current activity's "seasons per
country" bar chart, using `country2`:

```{r}
castaways |>
  count(state, sort = TRUE) |>
  head(10)
```

Heads up before building this into a question: `state` has
`r n_distinct(castaways$state, na.rm = TRUE)` distinct values (US states, plus
"state" fields from Survivor South Africa and Survivor AU, like Gauteng and
NSW) and 65 missing. A straight `geom_bar()` on all of it will be an unreadable
wall of one-off bars — which is honestly a decent motivating moment for "we'll
fix this exact problem with `forcats` in a few weeks," but not something to
solve in Round 5 itself. Cleanest fix for right now: `filter()` to the top ~10
states first (same move Round 5 already makes going from `country2` to
`country`), or just pick a handful of states to compare by name.

There's also a quirky, low-stakes variable here for the Round 1
`summary()`/`unique()` warm-up: `ack_score`, a score for how host Jeff Probst
acknowledges an eliminated player on the way out. It's sparse (about half
missing — looks like it's only tracked for more recent seasons) so it's not
plot-ready, but `unique(castaways$ack_score)` or a `summary()` makes for a fun
"wait, what is this column" moment before students even get to graphing.

## Option 3 (bonus): `confessionals` — screen time as data

```{r}
glimpse(confessionals)
```

`confessional_count` (how many confessionals a castaway gave in an episode) is
fully populated across `r n_distinct(confessionals$season)` seasons and ranges
0–22. Survivor superfans track this obsessively as a predictor of who's going
far ("edgic" theory) — genuinely fun hook if you have fans in the room.
`confessional_time` exists too but is missing for most rows (only recorded for
a subset of seasons), so lead with `confessional_count`.

```{r}
confessionals |>
  ggplot(aes(x = confessional_count)) +
  geom_histogram(binwidth = 1)
```

Right-skewed with a long tail — a nicer distribution-shape example than
anything in the current `season_summary`-based activity, since most of those
are close to symmetric.

## Making the plots more interesting, same skill level

A few `geom_*` additions that don't require anything beyond "you can add
another layer" — no new concepts, just slightly different visual payoff:

**`geom_smooth()`** — one extra line turns the "viewers over time" scatter
into an actual trend:

```{r}
episodes |>
  filter(!is.na(viewers)) |>
  ggplot(aes(x = episode_number_overall, y = viewers)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE)
```

**`geom_jitter()`** — already used in HW2, but worth pairing with a mostly-
discrete x-axis like `winner` or `jury_status` instead of a continuous one:

```{r}
castaways |>
  ggplot(aes(x = winner, y = day)) +
  geom_jitter(width = 0.15, alpha = 0.4)
```

**`geom_count()`** — same syntax as `geom_point()`, but sizes each point by
how many overlapping observations it represents. Nice for a scatter where both
variables are somewhat discrete (episode ratings, rounded to the nearest 0.1,
plotted against something else) — makes overplotting *informative* instead of
just messy, for zero extra conceptual cost.

None of these need facets, themes, or scales customization (still "next
week"), so they fit without pulling forward content from Slides 04.

## Bottom line

- **Best like-for-like upgrade:** `episodes` — same variables students already
  expect (`viewers`, `imdb_rating`), just far more of them, plus a real
  "expected a relationship, didn't find one" moment and a genuine trend to
  discover in Round 3.
- **Best change-of-pace:** `castaways` — moves from "seasons" to "people,"
  clean winner-vs-not comparison for the boxplot/violin round, and a bar-chart
  variable (`state`) that naturally previews the "too many categories" problem
  `forcats` solves later.
- **Best low-stakes bonus:** `confessionals` — one variable, well-populated,
  a nicer distribution shape, and a hook for anyone who's actually watched the
  show.

All three stay single-table, so nothing here requires pulling in Week 4's
joins content early.
