---
title: Lab Quiz 01 Info
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
library(tidyverse)
library(ggthemes)
library(nycflights23)
```

Our first lab quiz is scheduled for **Friday of Week 3**. It will only take ~40 minutes of the class period.

# Guidelines

This is a *closed note*, *closed internet resources*, *closed other people* lab quiz. I want to see what's in your brain! You may use the cheat sheets provided by me, but otherwise you may not use any resources.

The lab quizzes are not written to be tricky or very difficult. If you've been completing the in-class activities and the homework, and putting the time and effort in to understand them, you should do well on the lab quizzes.

# Format

The reason I give lab quizzes is because your life will be easier if you know how to do basic tasks in R "on the fly". I also want to see what you **know** and not just what you can do with access to your resources. However, I also know this is not how data science works in the real world, and so you can also use a token to complete a retake. 

1. The required in-class portion
  - Pencil-and-paper questions
  - Cheat sheets provided by me but no other resources
  - Designed to assess what you know "on the fly" without access to resources

2. An R-based retake
  - .qmd format similar to homework
  - Allowed to use materials from class (slides, activities, notes, homework, etc.) but no out-of-class resources (google, textbooks, ChatGPT or other LLMs, StackOverflow, etc.)
  - Designed to assess your coding skills and whether you can figure things out with access to class resources and R help pages



# Skills

## R Basics

-   Given a vector, list, or data frame, extract an element of interest.
    -   Know how to use each extractor (`x$__`, `x[__]`, `x[[__]]`) and what they return
    -   Construct logical vectors to use as an index to subset a vector or data frame
    -   Construct integer vectors to use as an index to subset a vector or data frame
-   Given a vector, list, or data frame, obtain a quick summary such as the length, dimension, or type of each element
    -   Functions to know: `length()`, `nrow()`, `ncol()`, `dim()`, `summary()`, `glimpse()`, `class()`, `typeof()`, `head()`, `tail()`.

## Data Visualization

-   Identify the appropriate layer to add to a static graphic in order to display specific information
    -   You should know the `geom` (and associated aesthetics) for the following charts: bar/column chart, histograms, boxplots, density plots, violin plots, scatterplots, time series line plots, map, chloropleth map
-   Given a graph and a dataset structure, fill in ggplot2 code to replicate the graphic:
    -   base layers
    -   axis labels, titles, captions
    -   `scale_x` functions for aesthetics
    -   `facet_wrap()` and `facet_grid()`
-   Given a question of interest and data set, decide on an appropriate graphic to address/answer the question
    -   You should know when the graphs mentioned above are appropriate
-   Given a graphic, identify which variables are shown and which aesthetic (x, y, color, size, etc.) each one is mapped to
-   Given a graphic, describe the strengths and weaknesses of it from a design perspective
    -   Core principles: show the data (don't distort it), choose the right plot, use color meaningfully and with restraint, tell a story, leave out non-story details
    -   Accessibility considerations: color, alt text, direct labeling, etc.

## Data Wrangling (covered in Week 3)

-   Know how the following verbs act on a data set: `filter`, `slice`, `select`
    -   Syntax for using
    -   Describe what the output would look like
-   Given a data set and goal, identify and utilize the appropriate verb to create the data set of interest
-   Use {dplyr} verbs AND the logical vector approach to subset a data frame 

# Grading

For the in-class portion, each question is worth 1 point, graded as "successful" (almost entirely correct), "not successful" (missing a key component), or "half credit" (mostly correct). 

For the out-of-class revision, each question is worth 1 point, graded as "successful" (correct), "not successful" (missing a key component), or "half credit" (mostly correct). Since you will be able to run the code, you will only receive full credit for fully correct solutions.

## Revision

[more details next week]

# Practice Questions

## Data

::: callout-tip

The quiz will use one dataset, and I will give you some information about it at the beginning of the quiz. All of the questions will use the same data, so if the answer to one question helps you with another, that's just fine!
:::

The `nycflights23` package contains information about all flights that departed from NYC (e.g. EWR, JFK and LGA) in 2023. The main data is in the `flights` data frame.

For this lab quiz, we're going to work with a random sample of 300 flights (called `flights_sampl`)

```{r}
set.seed(011725)
library(tidyverse)
library(nycflights23)
flights_sampl <- flights |>
  sample_n(300) |>
  select(year, month, day, carrier, origin, dest, sched_dep_time, dep_delay, air_time) |>
  left_join(airports |> select(dest = faa, lat, lon))

glimpse(flights_sampl)
```

## Questions

**Given a vector, list, or data frame, extract an element of interest.**

Write R code to do the following tasks: 

(a) Store the `dep_delay` column in a vector called `delay_minutes`

(b) Extract rows of `flights_sampl` where `dep_delay` is more than 15 minutes

::: callout-tip
The practice data here is a data frame. For practice with **lists** (including `[[ ]]` vs `[ ]`), see the [Lists](../notes/01-r-basics.qmd#lists) section of the R Basics reading and its check point.
:::


**Given a vector, list, or data frame, obtain a quick summary such as the length, dimension, or type of each element**

Carefully and clearly describe the result produced by the following commands: 

(a) `length(flights_sampl)`

(b) `dim(flights_sampl)`


**Identify the appropriate layer to add to a static graphic in order to display specific information**

Consider the `ggplot` code below.

```{r}
#| eval: false

ggplot(data = flights_sampl, aes(x = air_time, y = dep_delay)) + 
  geom_point() 
```

(a) Sketch a possible graph that this code would create

*Note:* I know that you do not know what the trends, etc. will look like, so you should focus on making sure that all of the aesthetics are correctly mapped and all of the layers are there

(b) If I wanted to color the points by month, how would I edit the code above?


**Given a graph and a dataset structure, provide ggplot2 code to replicate the graphic**

The map below displays the locations of airports with flights arriving from NYC. A few things to note: 

- The state borders are `black`
- The color within the borders is `lightblue`
- The map uses two datasets.
- The`states` dataset was generated with  `map_data("state")`. It has the following columns: 
    - `long`
    - `lat`
    - `group`
    - `order`
    - `region` 
- The second dataset is `flights_sampl`

```{r}
#| echo: false
states <- map_data("state")

ggplot(states) +
  geom_map(aes(map_id = region),
           map = states,
           color = "black", fill = "lightblue") +
  geom_point(data = flights_sampl,
             aes(x = lon, y = lat, col = carrier), size = 2) +
  expand_limits(x = states$long, y = states$lat) +
  coord_map() +
  theme_map() +
  theme(legend.position = "bottom") +
  scale_color_viridis_d(end = .75, option = "plasma")
```


Fill in the missing R code that is required to create this map:

```{r}
#| eval: false
states <- map_data("state")

ggplot(___) +
  geom_map(aes(map_id = ___),
           map = ___,
           ___) +
  geom_point(data = ___,
             aes(___),
             ___) +
  expand_limits(x = ___, y = ___) +
  coord_map() +
  theme_map() +
  theme(legend.position = "bottom") +
  scale_color_viridis_d(end = .75, option = "plasma")
```

**Given a question of interest and data set, decide on an appropriate graphic to address/answer the question**

Clearly describe one graph that you could make to answer the following questions: 

- Do flights in winter months have longer delays than summer months? 
- Is there a relationship between flight length and delay amount? 
- What carriers are most common at each of the three `origin` airports?


**Given a graphic, describe the strengths and weaknesses of it from a design perspective**

Consider the ggplot graphic below: 

```{r}
#| echo: false


flights_sampl |>
  group_by(dest) |>
  summarize(
    min = min(dep_delay, na.rm = T),
    max = max(dep_delay, na.rm = T),
    mean = mean(dep_delay, na.rm = T),
    n = n(),
  ) |>
  filter(n > 5) |>
  pivot_longer(-c(dest, n)) |>
  ggplot(aes(y = dest, x = value, fill = name)) +
  geom_col(position = "dodge") +
  labs(
    x = "Minutes",
    y = "Airport Code", 
    fill = ""
  ) +
  scale_fill_brewer(palette = "RdYlBu") + 
  theme_light() +
  theme(text = element_text(color = "lightblue"),
        axis.text = element_text(color = "lightblue"))
```

(a) Which variables are shown in this graph, and which aesthetic (e.g., x-axis, y-axis, color) is each one mapped to?
(b) Summarize what you think the "main point" of this graph is in 1 sentence
(c) Give one improvement this graph could make in the aesthetic or geometry layer to make the "main point" more understandable
(d) Give one improvement this graph could make in the theme layer to make it more readable or accessible

**Know how the following verbs act on a data set: `filter`, `slice`, `select`**

Describe or sketch out what the following R code would produce: 

```{r}
#| eval: false
flights_sampl |>
  filter(origin == "LGA") |>
  select(carrier, dep_delay)
```


**Given a data set and goal, identify and utilize the appropriate verb to create the data set of interest**

The small dataset below was generated from `flights_sampl`. What dplyr verbs were used? 

```{r}
#| echo: false

flights_sampl |>
  filter(dest == "MSP") |>
  select(month, dest, origin, air_time)
```


